---
title: "White Wine Quality Analysis"
author: "Hani Elmalky"
date: January 15, 2019
output: 
  html_document: 
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2 
    number_sections: yes
    theme: spacelab
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Installing all packages
#install.packages("ggplot2")
#install.packages("colorspace")
#install.packages("gridExtra")
#install.packages("dplyr")
#install.packages("magrittr")
#install.packages("GGally")
#install.packages("corrplot")
#install.packages("ggpubr")
#install.packages("RColorBrewer")
#install.packages("memisc")
#install.packages("cowplot")



# Loading all packages.

library(ggplot2)
library(gridExtra)
library(dplyr)
library(magrittr)
library(GGally)
library(corrplot)
library(ggpubr)
library(RColorBrewer)
library(memisc)
library(cowplot)


```

```{r echo=FALSE, Load_the_Data}
ww <- read.csv('wineQualityWhites.csv')
# Load the Data
```

# Abstract

The purpose of this analysis is to identify physical & chemical properties affecting white wine quality.

# Dataset

The dataset containing quality ranking of three wine tasting experts with details of chemical composition of 4898 white wine samples 

This dataset is made public by, 

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016 [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib


# Exploratory Analysis

## Dataset Structure Investigation

Dataset dimensions

```{r echo=FALSE, df_dim}
dim(ww)
```


Dataset content
```{r echo=FALSE, String}
str(ww)
```

The dataset includes 13 columns (1x index, 11x input variables, 1x output attribute)

We will drop index column `X`
```{r echo=FALSE, warning=FALSE, Summary}
ww$X <- NULL
```

New dataset dimensions
```{r echo=FALSE}
dim(ww)
INDEPENDENT <- colnames(ww)[1:11]

```

## Univariant Analysis
### Univariate Plots

Destiny plot of all variables & attributes 

```{r Distribution_only, eval=FALSE, fig.height=8, fig.width=12, warning=TRUE, include=FALSE}

plots <- list()
for (item in names(ww)) {
  plots[[item]] <- ggplot(data = ww, aes_string(x = item)) + 
    geom_density(fill = 'lightgray') + 
    geom_vline(aes(xintercept = mean(!!ensym(item))), linetype = "solid", color = 'blue', size = 0.6) + 
    geom_vline(aes(xintercept = median(!!ensym(item))), linetype = "solid", color = 'red', size = 0.6) + 
    geom_vline(aes(xintercept = quantile(!!ensym(item), probs = 0.25)), linetype = "dashed") +
    geom_vline(aes(xintercept = quantile(!!ensym(item), probs = 0.75)), linetype = "dashed") +
    labs( y="")
}

grid.arrange(grobs = plots, ncol = 3)

```

```{r Boxplots_only, eval=FALSE, fig.height=8, fig.width=12, warning=TRUE, include=FALSE}

plots <- list()

for (item in names(ww)) {
  plots[[item]] <- ggplot(data = ww) + 
    geom_boxplot(aes_string(y = item)) +
    theme(legend.position="none") +
    scale_fill_brewer(palette="Set1") +
    coord_flip()
}

grid.arrange(grobs = plots, ncol = 3)

```

```{r, fig.width=12, fig.height=16, echo=FALSE, warning=TRUE, Univariate_Plotmatrix}
plots <- list()
index <- 1
for (item in names(ww)) {
  if (!(index%%2) & !((index-1)%%3)) {
    index <- index + 3
  }
  plots[[index]] <- ggplot(data = ww, aes_string(x = item)) + 
    geom_density(fill = 'lightgray') + 
    geom_vline(aes(xintercept = mean(!!ensym(item))), linetype = "solid", color = 'blue', size = 0.6) + 
    geom_vline(aes(xintercept = median(!!ensym(item))), linetype = "solid", color = 'red', size = 0.6) + 
    geom_vline(aes(xintercept = quantile(!!ensym(item), probs = 0.25)), linetype = "dashed") +
    geom_vline(aes(xintercept = quantile(!!ensym(item), probs = 0.75)), linetype = "dashed") +
    labs( y="")
  
  
  plots[[index + 3]] <- ggplot(data = ww) + 
    geom_boxplot(aes_string(y = item)) +
    theme(legend.position="none") +
    scale_fill_brewer(palette="Set1") +
    coord_flip()
  
  index <- index + 1
}

grid.arrange(grobs = plots, ncol = 3)
```


Looking into each variable in isolation from other attributes.

### Analyzing Input Variables
```{r}
summary(subset(ww, select = -c(quality)))
```

   1 - **fixed acidity** (tartaric acid - $g / dm^3$) \
   Account for nonvolatile acids in wine, normally distributed around 6.8 $g/dm^3$
   
   2 - **volatile acidity** (acetic acid - $g / dm^3$) \
   Accounts of vinegar taste in wine and follow right skewed unimodal distribution 
   
   3 - **citric acid** ($g / dm^3$) \
   Adds 'freshness' to the wine, found in small quantities. Follows a right skewed unimodal distribution 
   
   4 - **residual sugar** ($g / dm^3$) \
   Right skewed distribution with high peak around the lower edge of IQR (at 1.7 $g/dm^3$)
   
   5 - **chlorides** (sodium chloride - $g / dm^3$) \
   Accounts for salty taste in wine, right skewed with 75% of sample below 0.05$g/dm^3$  
   
   6 - **free sulfur dioxide** ($mg / dm^3$) \
   Right skewed normal distribution around 34.00$mg/dm^3$
   
   7 - **total sulfur dioxide** ($mg / dm^3$) \
   Right skewed normal distribution around 134.0$mg/dm^3$
   
   8 - **density** ($g / cm^3$) \
   Density follow a normal distribution with mean at 0.994
   
   9 - **pH** (0 most acid, 7 neutral, 14 most base) \
   pH is a contentious scale represents acidity where 0 is the most acid, 7 is neutral, and 14 is the most base.
   Analysis shows a normal distribution of pH around 3.1 with IQR ~0.2
   
   10 - **sulphates** (potassium sulphate - $g / dm3$) \
   Sulphates is a wine additive acts as antimicrobial and antioxidant. Distribution shows a skew to the right of distribution.
   
   11 - **alcohol** (% by volume) \
   Alcohol follow a right skewed distribution with IQR between 9.50% and 11.40% by volume. yet the distribution shape is almost bi-modal (two peaks) with another peak around 12.5% by volume

### Analyzing Output Attributes
   12 - **quality** (score between 0 and 10) \
```{r, echo= FALSE}
summary(ww$quality)
```
   Quality distribution appears unimodal normal distribution centered at 6. with most wines have grades of 5, 6, and 7.
   
   Create a new categorical variable `quality.grp` from `quality` as per the following slices
    - Bad   : Wine quality rating less than or equal 5
    - Good  : Wine quality rating between 5 and 6 (6 included)
    - Great : Wine quality rating above 6
      
```{r, echo= FALSE}
ww$quality.grp <- cut(ww$quality, 
                      breaks = c(-Inf, 5, 6, Inf), 
                      labels = c("Bad", "Good", "Great"))

summary(ww$quality.grp)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ww, aes(x = quality, color = quality.grp, fill = quality.grp))+
  geom_jitter(aes(y = quality.grp), size = 0.6) + 
  scale_fill_brewer(palette="Set1")
```


### Univariant Analysis Summary

  - 75% of the dataset (by count) have *quality* value less than or equal to 6.00 (the median value)
    - Only 25% of samples in the dataset sample is classified as Great
  
  - *Alcohol*, *Ph*, *Total Sulfur Dioxide*, & *Sulphates* seams to have a large IQR compared to total count.
  
  - *Residual Sugar*, *Chlorides*, *Denisty*, *Free Sulfur Dioxide*, & *Citrix Acid* seams to have multiple outliers ($>2\mu$) in comparison to other input parameters.
  
  - A bivariant analysis is required to identify possible correlation across different parameters.

## Bivariant Analysis
### Correlation Analysis

We will start the bivariant analysis by identifying the correlation across different parameters in the data set.

```{r echo=FALSE, message=FALSE, warning=TRUE}

subset(ww, select = -c(quality.grp)) %>%
  as.matrix() %>%
  cor() -> corr_matrix
  
corrplot(corr_matrix , 
         method = c("pie"), 
         type = "lower", 
         diag = TRUE,
         tl.col = 'black', 
         tl.srt = 45)
```

Running the significance test at $\beta \leq 0.05$ and omit statistically insignificant results from the correlation graph.

```{r echo=FALSE, message=FALSE, warning=TRUE}

cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

plot_one <- corrplot(corr_matrix , 
                     method = c("pie"), 
                     type = "lower", 
                     diag = TRUE,
                     tl.col = 'black', 
                     tl.srt = 45, 
                     p.mat = cor.mtest(corr_matrix),
                     sig.level = 0.05, insig = "blank", addrect = 1)

```

Graph above infers the following significant correlation relations
  
  - Strong positive correlation between **Density** and **Residual Sugar**

  - Strong negative correlation between **Alcohol** and **Density**

  - Medium positive correlation between
  
      - **Quality** and **Alcohol**
      
      - **Density** and **Total Sulfur Dioxide** 
  
  - Medium negative correlation between
  
      - **Quality** and **Density**
      
      - **Alcohol** and each of (**Residual Sugar**, **Total Sulfur Dioxide**, **Chlorides**, **Free Sulfur Dioxide**)

\
The following table highlights correlation values between Quality & other variables in the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# create the correlation frame
corr_df <- subset(ww, select = -c(quality.grp)) %>%     # subset and remove categorial variable
  cor(., y = .$quality)  %>%                            # build correlation matrix
  round(., 3) %>%                                       # round all numbers to 2
  .[!rownames(.) %in% 'quality',] %>%                   # remove quality row from list
  as.data.frame(.) %>%                                  # build the list into a dataframe
  cbind(names = rownames(.), .)                         # copy row names into its column and extract

# clean row names & column names
rownames(corr_df) <- NULL
names(corr_df) <- c("Attributes", "QualityCorrelation")

# create absolute column
corr_df$AbsQualityCorrelation <- abs(corr_df$QualityCorrelation)

# create IQR column
corrBreaks <- c(0, 
                quantile(corr_df$AbsQualityCorrelation, 0.25), 
                quantile(corr_df$AbsQualityCorrelation, 0.50),
                quantile(corr_df$AbsQualityCorrelation, 0.75), 
                1)

corr_df$IQR = cut(corr_df$AbsQualityCorrelation, 
                                       breaks = corrBreaks, 
                                       labels = c('Q1', 'Q2', 'Q3', 'Q4'))

corr_df <- arrange(corr_df, desc(AbsQualityCorrelation))
corr_df

```

Based on above correlation analysis. 

  - **Citric Acid**, **Free Sulfur Dioxide**, **Sulphates** are not correlated with wine quality.
  - **Residual Sugar**, **pH**, **Fixed Acidity** are very weakly correlated with wine quality.
  - **Total Sulfur Dioxide**, and **Volatile Acidity** are relatively correlated with wine quality (still correlation is considered weak compared to correlation observed in Q4)
  - **Alcohol**, **Density**, **Chlorides** are strongly correlated with wine quality.
  
Hence, we will 

  - Drop **Citric Acid**, **Free Sulfur Dioxide**, **Sulphates**, **pH**, **Fixed Acidity** from the rest of this analysis.
  - Irrespective of weak correlation between Residual sugar & wine quality, We will keep **Residual Sugar** due to it's strong correlation with **Density**.
  - Focus the study on **Alcohol**, **Density**, **Chlorides**, **Total Sulfur Dioxide**, **Volatile Acidity**, **Residual Sugar** and their relationship with **Wine Quality**
  
  
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
ww$citric.acid <- NULL
ww$free.sulfur.dioxide <- NULL
ww$sulphates <- NULL
ww$pH <- NULL
ww$fixed.acidity <- NULL
```
  

### Bivariant Plots
We will start by plotting all variables against each others mapped to Wine Quality grade (Bad, Good, Great)
```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, retina=1}

#pairs(ww)

set.seed(19790427)
ww_sample <- ww[sample(1:length(ww$quality), 1000), ]

reg_2 <- function(data, mapping, ...){
  p <- ggplot(data = data, mapping = mapping) + 
    geom_point(alpha = 1/3, size = 1) + 
    geom_smooth(method=lm, fill="orange", color="darkgray", ...)
  p
}

ggpairs(ww_sample, axisLabels = 'show', 
        columns = c("quality", "alcohol", "density", "volatile.acidity", 
                    "chlorides", "total.sulfur.dioxide", "residual.sugar"), 
        legend = c(1,1),
        diag = list(mapping = aes(color = quality.grp)), 
        upper = list(continuous = wrap("cor", size=5, fontface = "bold")), 
        lower = list(continuous = reg_2, 
                     mapping = aes(color = quality.grp)
                     ),
        title = "Physical-Chemical Pairs-plot color coded by Wine Grade"
        ) + 
  theme(legend.position = "bottom", 
        panel.grid.major = element_blank(), 
        axis.ticks = element_blank(), 
        panel.border = element_rect(linetype = "solid", colour = "black", fill = NA))
# combo = wrap("dot", outlier.shape = I('.'))
```

We will use boxplot per wine grade to understand descriptional properties of different variables deeper.

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, retina=1}

plots <- list()
for (item in names(ww)[1:6]) {
  plots[[item]] <- ggplot(data = ww,
                          aes(fill = quality.grp)) + 
    geom_boxplot(aes_string(y = item)) +
    theme(legend.position="none") +
    scale_fill_brewer(palette="Set1") +
    coord_flip()
}

grid.arrange(grobs = plots, ncol = 3)

```

It is clear from the above box plot that we have outliers in multiple variables that may cloud our conclusions. Hence, we will be subsetting the dataset to reflect only 95% of each variable individually.

> We will NOT drop any data from the dataset at this point, instead, we will adjust analysis window for the 95 percentile to eliminate outliers noise.

In the next few sections, we will try to analyze Quality *as the main output variable* together with building a deeper understanding of other variables inter-dependencies.

#### Quality

Plotting Quality vs. other wine attributes while maintaining color-code for wine grade (Bad, Good, Great) to identify if there is any pattern associated with great wines.

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
corr_df2 <- subset(corr_df, !(corr_df[,1] %in% c("pH", "sulphates", "citric.acid", "free.sulfur.dioxide", "fixed.acidity")))
items <- corr_df2[,1]
plots <- list()


for (item in items) {
  plots[[item]] <- ggplot(data = ww, 
                          aes(x = quality, group = quality, fill = quality.grp)) + 
    geom_boxplot(aes_string(y = item)) +
    theme(legend.position="none") +
    scale_fill_brewer(palette="Set1") + 
    labs(fill = "Wine Grade") + 
    geom_smooth(method = "lm", se = TRUE, color = "orange", aes_string(y = item, group = 1))
}

grid.arrange(grobs = plots, ncol = 3)

#trial-code to add corr factor as text on the graph
  #corr_lable <- paste(c("r=", subset(corr_df2, (corr_df2[,1] %in% item))[,2]), collapse = " ")
  #print(corr_lable)

  #    +
  #geom_text(aes(x = 9, y = max(!!ensym(item)), 
  #               label = paste(c("r=", subset(corr_df2, (corr_df2[,1] %in% item))[,2]), collapse = " ")), 
  #          colour = 'black')

```

It is clear from the above box diagram that great quality wines have a strong positive correlation with alcohol density. The linear model plotted in orange shows a strong linear growth of wine quality with alcohol increase in wine.

The box plot also highlights the strong negative correlation between wine quality and Density, Chlorides and Volatile Acidity but the fact we have multiple outliers in those graphs is what is making the visual association harder.

Hence, we will plot the same parameters distribution per wine grade for ascending 95% ~ 98% of the dataset population to eliminate the last 2% ~ 5% outliers.

  **Density distribution of quality key paraemters, color coded by wine grade**

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
p1 <- ggdensity(ww, 
          x = "alcohol", color = "quality.grp", fill="quality.grp", 
          add = "mean", rug = FALSE,  palette = "Set1")

p2 <- ggdensity(ww, 
          x = "density", color = "quality.grp", fill="quality.grp", 
          add = "mean", rug = FALSE,  palette = "Set1") %>%
  ggpar(., xlim = c(min(ww$density), quantile(ww$density, probs = 0.99)))

p3 <- ggdensity(ww, 
          x = "chlorides", color = "quality.grp", fill="quality.grp", 
          add = "mean", rug = FALSE,  palette = "Set1") %>%
  ggpar(., xlim = c(min(ww$chlorides), quantile(ww$chlorides, probs = 0.95)))

p4 <- ggdensity(ww, 
          x = "volatile.acidity", color = "quality.grp", fill="quality.grp", 
          add = "mean", rug = FALSE,  palette = "Set1") %>%
  ggpar(., xlim = c(min(ww$volatile.acidity), quantile(ww$volatile.acidity, probs = 0.95)))

ggarrange(p1, p2, p3, p4, ncol=2, nrow = 2, common.legend = TRUE, legend="bottom", labels = "AUTO")

```
  
  We can infer from the above plot that, 
  
  A. Great quality wines have the highest median Alcohol level.
  
  B. Great quality wines have the lowest median Density.
  
  C. Great quality wines tend to have less Chlorides.
  
  D. Great quality wines tend to have less Volatile Acidity levels.
  
#### Alcohol

```{r echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
items <- c("density", "residual.sugar", "total.sulfur.dioxide", "chlorides")
color_series <- c("quality.grp")
plots <- list()

for (item in items) {
  plots[[item]] <- ggplot(data = ww, aes(x = alcohol)) + 
    geom_point(alpha = 1/5,aes_string(y = item, colour = item)) +
    scale_colour_gradientn(colours=rainbow(4))+
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1))
}

grid.arrange(grobs = plots, ncol = 2)
```

Alcohol have a strong negative correlation with **Density**, and a weak negative correlation with **Residual Sugar**, **Total Sulfur Dioxide**, and **Chlorides**.

#### Density

> Dataset used in the following plot has been modified to remove outliers above the 95 percentile of the under-analysis attributes.

```{r echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
items <- c("residual.sugar", "total.sulfur.dioxide")
x_item <- c("density")
plots <- list()

ww_subset <- subset(ww, ww[[x_item]] <= quantile(ww[[x_item]], 0.95))

for (item in items) {
  plots[[item]] <- ggplot(data = subset(ww_subset, ww_subset[[item]] <= quantile(ww_subset[[item]], 0.95)), 
                          aes_string(x = x_item)) + 
    geom_point(alpha = 1/5,aes_string(y = item, colour = item)) +
    scale_colour_gradientn(colours=rainbow(4))+
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    theme(legend.position="bottom")
}

grid.arrange(grobs = plots, ncol = 2)
```

Analysis shows that 

- **Density** is strongly correlated with **Residual Sugar** at r =  **`r round(cor(ww$density, ww$residual.sugar),2)`**

- **Density** is weakly correlated with **Total Sulfur Dioxide** at r = **`r round(cor(ww$density, ww$total.sulfur.dioxide),2)`**

#### Residual Sugar & Total Sulfur Dioxide

> Dataset used in the following plot has been modified to remove outliers above the 95 percentile of the under-analysis attributes.

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
items <- c("total.sulfur.dioxide")
x_item <- c("residual.sugar")
plots <- list()

ww_subset <- subset(ww, ww[[x_item]] <= quantile(ww[[x_item]], 0.95))

for (item in items) {
  plots[[item]] <- ggplot(data = subset(ww_subset, ww_subset[[item]] <= quantile(ww_subset[[item]], 0.95)), 
                          aes_string(x = x_item)) + 
    geom_point(alpha = 1/5,aes_string(y = item, colour = item)) +
    scale_colour_gradientn(colours=rainbow(4))+
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1))
}

grid.arrange(grobs = plots, ncol = 1)

```

Previous plot shows a weak correlation between **Residual Sugar** and **Total Sulfur Dioxide** with correlation coefficient "r" equals **`r round(cor(ww$residual.sugar, ww$total.sulfur.dioxide),2)`**


### Bivariant Analysis Summary

With regards to Wine Quality
 
- Great quality wines have the highest median **Alcohol** levels and the lowest median **Density** levels.

- Great quality wines tend to have less **Chlorides** & **Volatile Acidity** levels.

\
We have also noticed the following strong correlations in the dataset

- **Alchohol** is strongly correlated with **Density** with correlation coefficient "r" equals  **`r round(cor(ww$alcohol, ww$density),2)`**.

- **Density** is strongly correlated with **Residual Sugar** with correlation coefficient "r" equals  **`r round(cor(ww$density, ww$residual.sugar),2)`**.


## Multivariate Analysis

### Multivariant Plots

> Dataset used in the following plot has been modified to remove outliers above the 95 percentile of the **Density** attributes.


#### Quality interdependency on Alcohol vs. Density

Based on the findings in the bivariant analysis, in this section we will try to understand the relationship between wine quality and its strongly correlated variables in a multivariate environment.

```{r echo=FALSE, fig.height=8, fig.width=12}

items <- c("alcohol")
x_item <- c("density")
color_item1 <- ("quality.grp")
color_item2 <- ("quality")
plots <- list()
index <- 1

ww_subset <- subset(ww, ww$density <= quantile(ww$density, 0.95))

for (item in items) {
  plots[[index]] <- ggplot(data = ww_subset, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item1)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_fill_brewer(palette="Set1") + 
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)


  
  plots[[index + 1]] <- ggplot(data = ww_subset, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item2)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_colour_gradientn(colours=rainbow(4)) +
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)
    # scale_colour_gradient2(low = "white", mid = "green" , high = "red")

  index <- index + 2
}

grid.arrange(grobs = plots, ncol = 2, bottom = "dashed lines represent the mean of its corresponding axis")

```

Above diagram infer that, for the same **Density** level, great quality wines tend to have higher **Alcohol** content.

Let's isolate the great-quality wine in its own graph to see if this statement still hold.


```{r echo=FALSE, fig.height=8, fig.width=12}

ggplot(data = ww_subset, aes(x = density)) + 
  geom_point(alpha = 1,aes(y = alcohol, colour = quality)) +
  geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes(y = alcohol, group = 1)) +
  facet_wrap(~quality.grp, ncol = 3) + 
  scale_colour_gradientn(colours=rainbow(4)) + 
  geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
  geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)


```

As we can see in the above plot, the great wine quality category tend to have higher rating with increased alcohol levels at the same density range.

On the other hand, the Bad wine quality category tend to have very low ratings with increased alcohol levels. Looks like there is a second order variable in play here.

Hence, we will start analyzing those Quality vs. second order variables.

#### Quality interdependency on Second order variables vs. Alcohol

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}

items <- c("residual.sugar", "chlorides", "volatile.acidity")
x_item <- c("alcohol")
color_item1 <- ("quality.grp")
color_item2 <- ("quality")
plots <- list()
index <- 1


for (item in items) {
  plots[[index]] <- ggplot(data = ww_subset, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item1)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_fill_brewer(palette="Set1")
  
  plots[[index + 1]] <- ggplot(data = ww_subset, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item2)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_colour_gradientn(colours=rainbow(4))
    # scale_colour_gradient2(low = "white", mid = "green" , high = "red")

  index <- index + 2
}

grid.arrange(grobs = plots, ncol = 2)

```
We can infer from the above diagram that. 

- For the same **Alcohol** content, great quality wines tend to have lower **Residual Sugar** range.
- Increased levels of **Chlorides** happens more with lower grades of wine.

We can increase the plot contrast via dropping the middle grade (Good Wine) wine group from the dataset & try to replot the graphs one more time.

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}

items <- c("residual.sugar", "chlorides", "volatile.acidity")
x_item <- c("alcohol")
color_item1 <- ("quality.grp")
color_item2 <- ("quality")
plot_three <- list()
index <- 1

ww_subset_bad_great <- subset(ww, ww$quality.grp == c("Bad", "Great"))

for (item in items) {
  plot_three[[index]] <- ggplot(data = ww_subset_bad_great, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item1)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_fill_brewer(palette="Set1") + 
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)


  
  plot_three[[index + 1]] <- ggplot(data = ww_subset_bad_great, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item2)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_colour_gradientn(colours=rainbow(4)) +
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)
    # scale_colour_gradient2(low = "white", mid = "green" , high = "red")

  index <- index + 2
}

grid.arrange(grobs = plot_three, ncol = 2, bottom = "dashed lines represent the mean of its corresponding axis")
```

It is clear from the above plot that, 

- For the same **Alcohol** content, great quality wines tend to have lower **Residual Sugar** value & range.
- High levels of chlorides only exists in bad quality wines.
- For the same **Volatile Acidity** level, great quality wines tend to have higher **Alcohol** content.
- Samples with high **Volatile Acidity** or high **Residual Sugar** tend to have lower than average **Alcohol** levels.


#### Quality interdependency on Second order variables vs. Density

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}

items <- c("residual.sugar", "total.sulfur.dioxide", "volatile.acidity")
x_item <- c("density")
color_item1 <- ("quality.grp")
color_item2 <- ("quality")
plots <- list()
index <- 1


for (item in items) {
  plots[[index]] <- ggplot(data = ww_subset, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item1)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_fill_brewer(palette="Set1")
  
  plots[[index + 1]] <- ggplot(data = ww_subset, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item2)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_colour_gradientn(colours=rainbow(4))
    # scale_colour_gradient2(low = "white", mid = "green" , high = "red")

  index <- index + 2
}

grid.arrange(grobs = plots, ncol = 2)

```

It is hard to draw conclusions from the above plot, let's try to drop the middle grade wine (Good Wine) from the data set and try again.

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}

items <- c("residual.sugar", "total.sulfur.dioxide", "volatile.acidity")
x_item <- c("density")
color_item1 <- ("quality.grp")
color_item2 <- ("quality")
plots <- list()
index <- 1

ww_subset_bad_great <- subset(ww, ww$quality.grp == c("Bad", "Great"))

for (item in items) {
  plots[[index]] <- ggplot(data = ww_subset_bad_great, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item1)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_fill_brewer(palette="Set1") + 
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)
  
  plots[[index + 1]] <- ggplot(data = ww_subset_bad_great, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item2)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_colour_gradientn(colours=rainbow(4)) +
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)
    # scale_colour_gradient2(low = "white", mid = "green" , high = "red")

  index <- index + 2
}

grid.arrange(grobs = plots, ncol = 2, bottom = "dashed lines represent the mean of its corresponding axis")

```

We can infer from the above plot that, 

- Great quality wines tend to have lower **Density**.
- For the same **Density** level, great quality wines tend to have higher **Residual Sugar** value.
- Great quality wines tend to have a lower **Total Sulfur Dioxide** range compared to bad quality wines. 

#### Quality interdependency Residual Sugar & Total Sulfur Dioxide

> Dataset used in the following plot has been modified to remove outliers above the 95 percentile of the **Total Sulfur Dioxide** & **Residual Sugar** attributes.

> Dataset used in the following plot has been modified to remove the **Good** quality subgroup for better contrast.


```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}

items <- c("total.sulfur.dioxide")
x_item <- c("residual.sugar")
color_item1 <- ("quality.grp")
color_item2 <- ("quality")
plots <- list()
index <- 1

ww_subset2 <- subset(ww, 
                     ww$total.sulfur.dioxide <= quantile(ww$total.sulfur.dioxide, 0.95) & 
                       ww$residual.sugar <= quantile(ww$residual.sugar, 0.95) & 
                       ww$quality.grp == c("Bad" ,"Great"))

for (item in items) {
  plots[[index]] <- ggplot(data = ww_subset2, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item1)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_fill_brewer(palette="Set1") + geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)
  
  plots[[index + 1]] <- ggplot(data = ww_subset2, aes_string(x = x_item)) + 
    geom_point(alpha = 1, size = 1,aes_string(y = item, colour = color_item2)) +
    geom_smooth(method = "lm", se = TRUE, color = "orange", fill="blue", aes_string(y = item, group = 1)) +
    scale_colour_gradientn(colours=rainbow(4)) + 
    geom_vline(aes(xintercept = mean(!!ensym(x_item))), linetype = "dashed", color = 'darkgray', size = 0.6) +
    geom_hline(aes(yintercept = mean(!!ensym(item))), linetype = "dashed", color = 'darkgray', size = 0.6)
    # scale_colour_gradient2(low = "white", mid = "green" , high = "red")

  index <- index + 2
}

grid.arrange(grobs = plots, ncol = 2, bottom = "dashed lines represent the mean of its corresponding axis")

```

We can infer from the above plot that, For the same **Residual Sugar** level, great quality wines tend to have lower **Total Sulfur Dioxide** content.

### Multivariate Analysis Summary

We can summarize the key conclusions we gathered from the multivariate analysis into the following list.

- Higher grade wines are the ones with the highest level of **Alcohol** & lowest **Density**.
- For the same **Alcohol** content, great quality wines tend to have lower value & range of **Residual Sugar**.
- For the same **Volatile Acidity** level, great quality wines tend to have higher **Alcohol** content.
- For the same **Density** level, higher quality wine tend to have higher **Alcohol** content.
- For the same **Density** level, great quality wines tend to have higher **Residual Sugar** value.
- For the same **Residual Sugar** level, great quality wines tend to have lower **Total Sulfur Dioxide** content.
- High levels of chlorides only exists in bad quality wines.
- Samples with high **Volatile Acidity** or high **Residual Sugar** tend to have lower than average **Alcohol** levels.


# Linear Regression

```{r}
## Load fresh version of data
ww_lm <- read.csv('wineQualityWhites.csv')

m1 <- lm(quality ~ alcohol, data = ww_lm)
m2 <- update(m1, ~ . - alcohol + density)
m3 <- update(m2, ~ . + alcohol)
m4 <- update(m3, ~ . + fixed.acidity + volatile.acidity + citric.acid + residual.sugar + 
               chlorides + free.sulfur.dioxide + total.sulfur.dioxide + pH + sulphates)

mtable(m1, m2, m3, m4, sdigits = 3)

# fml1 <- as.formula(paste("quality", "~", paste(INDEPENDENT, collapse=' + ')))
#m1 <- lm(fml1, ww_lm)
```


Linear regression analysis has led to the following observations, 

- ~**20%** of **Quality** variance is explained by change in **Alcohol** level.
- ~**1%** of **Quality** variance is explained by change in **Density** level.
- ~**28%** of **Quality** variance is explained by change in all independent variables in the dataset.

**Conclusion:**

  Linear modeling is not sufficient to predict white wine quality.

# Final Plots and Summary

### White Whine physical & chemical attributes correlations
```{r echo=FALSE, fig.height=6, fig.width=9}
corrplot(plot_one , 
         method = c("pie"), 
         type = "lower", 
         diag = TRUE,
         tl.col = 'black', 
         tl.srt = 0.001,
         p.mat = cor.mtest(plot_one),
         title='Correlation of Phsyical & Chemical Attributes of White Wine', 
         sig.level = 0.05, insig = "blank", addrect = 1, 
         mar=c(0,0,1,0)
         )
```

Above correlation plot omits statistically insignificant results from the correlation graph at $\beta \leq 0.05$

Graph above infers the following significant correlation relations
  
  - Strong positive correlation between **Density** and **Residual Sugar**

  - Strong negative correlation between **Alcohol** and **Density**

  - Medium positive correlation between
  
      - **Quality** and **Alcohol**
      
      - **Density** and **Total Sulfur Dioxide** 
  
  - Medium negative correlation between
  
      - **Quality** and **Density**
      
      - **Alcohol** and each of (**Residual Sugar**, **Total Sulfur Dioxide**, **Chlorides**, **Free Sulfur Dioxide**)

### Quality vs. Alcohol & Density


**Density distribution of Alcohol & Density color coded by wine grade**

```{r echo=FALSE, fig.height=6, fig.width=12, message=FALSE, warning=FALSE, Plot_Two}
p1 <- ggdensity(ww, 
          x = "alcohol", color = "quality.grp", fill="quality.grp", 
          add = "mean", rug = FALSE,  palette = "Set1")

p2 <- ggdensity(ww, 
          x = "density", color = "quality.grp", fill="quality.grp", 
          add = "mean", rug = FALSE,  palette = "Set1") %>%
  ggpar(., xlim = c(min(ww$density), quantile(ww$density, probs = 0.99)))

ggarrange(p1, p2, ncol=2, nrow = 1, common.legend = TRUE, legend="bottom", labels = "AUTO")
```

Above plot infer that great quality wines have the highest median Alcohol level and the lowest median density.

### Quality interdependency on second order variables vs. Alchohol
```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, Plot_Three}
grid.arrange(grobs = plot_three, ncol = 2, bottom = "dashed lines represent the mean of its corresponding axis")

```

Above plot infers that

- For the same **Alcohol** content, great quality wines tend to have lower **Residual Sugar** value & range.

- High levels of chlorides only exists in bad quality wines.

- For the same **Volatile Acidity** level, great quality wines tend to have higher **Alcohol** content.

- Samples with high **Volatile Acidity** or high **Residual Sugar** tend to have lower than average **Alcohol** levels.


------

# Reflection

The white wine dataset used in this analysis contains multiple physical & chemical attributes & properties together with rating of wine quality. across the 5000 samples, linear regression has failed to predict white wine quality. Yet we have noticed a strong correlation between white wine quality and alcohol where 20% of quality variance can be explained by change in alcohol content.

